---
title: "ccRCC ECM multi-cohort analysis"
author: "Ronen Levy"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

## Set TRUE to force every stage to recompute from source data, ignoring any
## existing checkpoint. FALSE (default) reuses a stage's checkpoint when
## present and only (re)runs a stage whose terminal checkpoint is missing.
run_from_scratch <- FALSE

repo <- normalizePath(".")
readRenviron(file.path(repo, ".Renviron"))
source(file.path(repo, "config.R"))

library(yaml)
params <- yaml::read_yaml(file.path(repo, "params.yaml"))

dir.create(file.path(repo, "checkpoints"), showWarnings = FALSE, recursive = TRUE)

#' Runs a stage's script sequence (each sourced via Rscript, so each script's
#' own commandArgs()-based repo-root detection still works unchanged) unless
#' `checkpoint_path` already exists and run_from_scratch is FALSE. This is
#' the ONLY orchestration logic added here - every analytical step is the
#' existing, already-validated script in R/, run unmodified via source().
run_stage <- function(stage_name, scripts, checkpoint_path, run_from_scratch = FALSE) {
  checkpoint_exists <- file.exists(checkpoint_path)
  if (!run_from_scratch && checkpoint_exists) {
    cat(sprintf("[%s] checkpoint found (%s) - skipping, reusing existing outputs.\n", stage_name, checkpoint_path))
    return(invisible(NULL))
  }
  cat(sprintf("[%s] checkpoint missing or run_from_scratch=TRUE - running %d script(s)...\n", stage_name, length(scripts)))
  for (s in scripts) {
    cat(sprintf("  source(R/%s)\n", s))
    source(file.path(repo, "R", s), local = new.env())
  }
  if (!file.exists(checkpoint_path)) {
    warning(sprintf("[%s] finished but expected checkpoint still missing: %s", stage_name, checkpoint_path))
  } else {
    cat(sprintf("[%s] done.\n", stage_name))
  }
}
```

Specification: `ECM_ccRCC_report_v4_tracked.docx` is the report this notebook
reproduces. Thresholds and settings are read from `params.yaml` (not
hard-coded below). See `README.md` for the full Records index and for what
is intentionally excluded (`exploratory/`, and Harmony/C-reference-mapping,
not included in this repository at all).

# 01 Load + QC

Builds each cohort's raw Seurat checkpoint from source files, characterizes
raw per-nucleus/per-cell QC, derives cohort C's %MT/non-MT-UMI thresholds,
applies the per-cohort QC filter (`params.yaml$qc`), collapses duplicate
gene symbols (A, C) onto a shared gene-identifier space, normalizes each
cohort, and runs the population-retention and 5%-MT-sensitivity checks plus
the combined QC figure (Fig. 1, Table 1).

```{r stage-01-load-qc}
run_stage(
  "01-load-qc",
  scripts = c(
    "run_load_cohort_a.R", "build_seurat_a.R",
    "build_seurat_b.R",
    "build_seurat_c.R",
    "qc_cohort_a.R", "qc_cohort_b.R", "qc_cohort_c.R",
    "derive_c_mt_threshold.R",
    "filter_cohort_a.R", "filter_cohort_b.R", "filter_cohort_c.R",
    "collapse_duplicate_symbols.R",
    "recollapse_cohort_c.R",
    "preprocess_cohort_a.R", "preprocess_cohort_b.R", "preprocess_cohort_c.R",
    "qc_population_retention.R", "qc_population_retention_A5pct.R",
    "qc_combined_ABC_figure.R"
  ),
  checkpoint_path = file.path(SCRATCH_DIR, "seurat_C_normalized.rds"),
  run_from_scratch = run_from_scratch
)
```

# 02 Integration

2,000 integration features selected jointly across A/B/C; RPCA anchor-based
joint integration (30 dims, k.weight 30); clustering and UMAP on 10 PCs.
Produces the unintegrated-baseline and RPCA-integrated UMAPs (Fig. 2a-b).
Clustering is swept at resolution 0.3, 0.5 and 1.0 for simplicity (kept
together in one stage); only resolution=1 is used downstream (Section 03,
cohort C labeling) and reported in the text - 0.3/0.5 are included as
supporting context for that choice, not separately reported.

```{r stage-02-integration}
run_stage(
  "02-integration",
  scripts = c(
    "prepare_integration_abc.R",
    "baseline_abc_umap.R",
    "prepare_rpca_pca_abc.R",
    "find_rpca_anchors_abc.R",
    "integrate_rpca_data_abc.R",
    "rpca_pca_only_abc.R",
    "rpca_umap_only_abc.R",
    "rpca_clusters_umap_abc.R",   # resolution 0.5 and 1.0
    "rpca_cluster_res03_abc.R",   # adds resolution 0.3 to the same sweep
    "crosstab_res03_population_abc.R"  # resolution 0.3 purity, for context only
  ),
  checkpoint_path = file.path(repo, "results", "tables", "integration", "ABC_res0.3_cluster_purity.tsv"),
  run_from_scratch = run_from_scratch
)
```

# 03 Cohort C labeling

A/B populations come from the harmonized `label_mapping.tsv`. Each C cell
receives the majority A/B population of its resolution=1 cluster when that
label covers >=80% of >=50 A/B cells, else Unresolved. Labels are checked
against canonical markers outside the ECM panel (Fig. 2c).

```{r stage-03-c-labeling}
run_stage(
  "03-c-labeling",
  scripts = c(
    "annotate_c_by_cluster_majority.R",
    "verify_annotation_markers.R",
    "plot_abc_umap_by_population.R"
  ),
  checkpoint_path = file.path(SCRATCH_DIR, "merged_ABC_annotated.rds"),
  run_from_scratch = run_from_scratch
)
```

# 04 ECM overview + population choice

ECM-panel expression across major populations - pooled, per-cohort (Fig. 3a)
and per-condition - plus the paired-donor coverage (>=10, >=20 cells in both
Tumor and Adjacent) per population x cohort that motivates choosing
Endothelial as the only population with consistent identity and paired
coverage in all three cohorts.

```{r stage-04-ecm-overview-population-choice}
run_stage(
  "04-ecm-overview-population-choice",
  scripts = c(
    "ecm_panel_by_population.R",
    "ecm_panel_by_population_per_cohort.R",
    "ecm_panel_by_population_per_condition.R",
    "ecm_population_choice_coverage.R"
  ),
  checkpoint_path = file.path(repo, "results", "ecm_population_choice", "paired_coverage_ge20.tsv"),
  run_from_scratch = run_from_scratch
)
```

# 05 Pseudobulk models + robustness

Raw UMI counts summed per donor x condition within Endothelial cells
(pseudobulk, >=20 cells/sample), TMM-normalized (edgeR) and modeled with
limma-voom (cohort as covariate, donor as a duplicateCorrelation block; a
strictly paired `~donor + condition` model on the 13 paired donors) - A
only, B only, C only, A+B, A+B+C, paired-only (Table 2). Robustness check:
Endothelial-labeled cells carrying any count of a mesenchymal marker
(`params.yaml$mesenchymal_markers`) are removed and every model refitted
unchanged.

```{r stage-05-pseudobulk-models-robustness}
run_stage(
  "05-pseudobulk-models-robustness",
  scripts = c(
    "ecm_endothelial_pseudobulk_build.R",
    "ecm_endothelial_models.R",
    "ecm_endothelial_mesenchymal_flag.R",
    "ecm_endothelial_clean_models.R",
    "ecm_endothelial_add_Aonly_clean.R"
  ),
  checkpoint_path = file.path(repo, "results", "ecm_endothelial", "clean", "before_after_comparison.tsv"),
  run_from_scratch = run_from_scratch
)
```

# 06 Linked modalities (cohort A)

Validates donor/specimen/aliquot correspondence against the supplied
multimodal crosswalk (cohort A only - the only cohort with linked bulk
RNA/proteomics) and compares the Tumor-vs-Adjacent direction of the chosen
ECM finding (Endothelial SPARC/FN1) against linked bulk RNA and protein,
within donor, per assay (Table 3).

```{r stage-06-linked-modalities}
run_stage(
  "06-linked-modalities",
  scripts = c(
    "multimodal_matched_subset.R",
    "multimodal_ecm_direction_check.R"
  ),
  checkpoint_path = file.path(repo, "results", "multimodal", "ecm_direction_paired_summary.tsv"),
  run_from_scratch = run_from_scratch
)
```

# Session info

```{r session-info}
sessionInfo()
```
